|

### 2023-02-05 manuscript/manuscript.Rmd
goodnight

In medical intervention research including in particular randomized controlled trials (RCTs) there is a trend towards increased use of patient reported outcomes; Quality of Life scores being one of the most prominent of these. Established statistical practice is to compare these between treatment groups using non-parametric tests such as Mann–Whitney–Wilcoxon rank-sum test (@wilcoxon3001968individual, @mann1947test), henceforth Wilcoxon, since distributions are rarely normal or symmetric. When all patients are alive and able and willing to provide quality of life scores at the scheduled measurement time this procedure works very well. Not all clinical settings, however, will have all participants alive at time of scheduled assessment of quality of life This is particularly true in trials within intensive care where mortality of approximately 30% is not uncommon and which is the setting that motivated this work. A pragmatic solution which is gaining popularity is to simply give the deceased patients the lowest possible score and then use a Wilcoxon test or similar to compare groups. This paper will demonstrate that this approach can lead to dramatic loss of statistical power. The paper will also present an alternative and novel statistical test that avoids the power loss. The novel test procedure is implemented in R and made publicly available.

>>>>

In medical intervention research including in particular randomized controlled trials (RCTs) there is a trend towards increased use of patient reported outcomes; Quality of Life scores being one of the most prominent of these. Established statistical practice is to compare these between treatment groups using non-parametric tests such as Mann–Whitney–Wilcoxon rank-sum test (@wilcoxon3001968individual, @mann1947test), henceforth Wilcoxon, since distributions are rarely normal or symmetric \textbf{[TODO: References needed to current statistical practice]}. When all patients are alive and able and willing to provide quality of life scores at the scheduled measurement time this procedure works very well. Not all clinical settings, however, will have all participants alive at time of scheduled assessment of quality of life This is particularly true in trials within intensive care where mortality of approximately 30% is not uncommon. This is the setting that motivated this work. A pragmatic solution which is gaining popularity is to simply give the deceased patients the lowest possible score and then use a Wilcoxon test or similar to compare the groups \textbf{[TODO: References needed]}. This paper demonstrates that this approach can lead to a dramatic loss of statistical power, and we will introduce an alternative statistical test statistic that averts this power loss. Our proposed test procedure is implemented in an R package and made publicly available.
